Interpretable Neural Networks using EAGGA

### 2025-03-14 paper/main.tex
add NN architecture + dataset split visualisation

(Taxonomy)
\citet[chap. 2]{survey_NN_interpretability} extend this distinction to a 3-dimensional taxonomy, allowing for better categorisation of neural networks (NN), a model class
that in its fully-connected feedforwad form is inherently non-interpretable.
% alternative
%As fully-connected feedforward neural networks (NNs) are inherently non-interpretable models, the taxonomy introduced by \citet[chap. 2]{survey_NN_interpretability} proves
%particulary useful to categorise NN explanability methods along 3 dimensions:
- \textbf{Passive vs Active Approaches}
  * passive: post-hoc
  * active: changing architecture or training process to make model interpretable
- \textbf{Type of Explanations}
  * examples: providing examples of what leads to desired output
  * attribution: attributing effect on output for a specific feature
  * hidden semantics: examine what type of inputs specific neurons / layers pick up on
  * rules: logical rules, e.g. if-then, trees
- \textbf{Local vs Global Interpretability}
  * local: explanation based on individual samples
  * semi-local: explanation based on set of samples, e.g. grouped together by some criterion
  * global: explaining network as a whole

(Evaluation)
- objective evaluation difficult, as no clear definition of interpretability
- \citet[3]{DoshiVelez2017TowardsAR} propose a taxonomy to categorise possible evaluation methods
  * \textbf{application-grounded evaluation}
    - most rigorous method, also most expensive + time consuming
    - idea: evaluate ``interpretability model'' directly w.r.t. the task -> human experts evaluate the outcome
    - e.g.: model performing medical diagnosis on patients would be evaluated by based on doctors doing the same
  * \textbf{human-grounded metrics}
    - similar to application-grounded, but tries to simplify task (while preserving its essence) so that laypersons can do it
    - cheaper due to larger, less qualified subject pool
    - especially suitable if only general concepts of the tasks need to be validated
    - examples for evaluation set-up
      * binary forced choice: human evaluator chooses which of two explanations he finds better
      * forward simulation: evaluator must correctly simulate model output when presented model input + explanation % perhaps binary forced choice is already enough to include, this (and the third one is a bit awkward to explain why this is good for interpretability imo)
  * \textbf{functionally-grounded evaluation}
    - assess explanation quality via some formally defined proxy for interpretability
    - no human time nor cost required beyond initial formulation of proxy
    - e.g. if we already have an interpretable model class (e.g. identified via human-grounded evaluation), can then rank different models of that class based on proxy
    - main challenge: what proxies

@@

(evolutionary algorithms)
- another model-free class of optimisation algorithms, inspired by naturally occuring evolution
- based on a population, which iteratively (iteration = generation) generates $\lambda$ offspring via 3 operators (all sub-bullets from \citet[pp. 10-]{genetic_algos})
  * \textbf{reproduction}: pick an individual to reproduce with p proportional to its fitness
  * \textbf{crossover}
    - select 2 ``parents'' from pool of reproducing individuals
    - for a position $k$ in their HP configs, with certain p, swap their values after $k$, yielding 2 children
    %use example from book page 12
  * \textbf{mutation}: with certain p, change values of an individual, e.g. by adding Gaussian noise (Gaussian mutation) to real values, or flipping bit for binary values
- selection at end of each genertion: keep $\mu$ best individuals (based on some fitness function), either from only offspring (``$(\mu,\lambda)$-selection'')
  or (more commonly) from population + offspring (``$(\mu+\lambda)$-selection'', guaranteed to keep best individual)
- pro: conceptually simple + can handle even complex parameter spaces (continuous, discrete, hierarchical, etc.) given appropriate implementation of operators
all above starting at HPO from \citep[chap. 1.3]{feurer_hyperparameter_2019}
- prominent examples
  * CMA-ES
    - ``Covariance Matrix Adapation Evolution Strategy''
    - offspring generation exclusively via multivariate normal \citep[p. 8]{hansen2023cmaevolutionstrategytutorial}
      * mean = weighted average of previous generation
      * covariance = weighted covariance of previous generation, with weighting as for mean
    - weighing scheme done to sample in a way as to reproduce previously successful (i.e. selected) steps \citep[p. 11]{hansen2023cmaevolutionstrategytutorial}
  * differential evolution \citep[-]{differential_evolution}
    - init population randomly so that entire param space is covered
      * $D$-dim vectors $\boldsymbol{x}_{i,G}$ with $i=1,2,...,n$ for gen $G$
    - mutation
      * for each $\boldsymbol{x}_{i,G}$, generate \textbf{mutant vector} $\boldsymbol{v}_{i,G+1}=\boldsymbol{x}_{r_1,G}+F\cdot(\boldsymbol{x}_{r_2,G}-\boldsymbol{x}_{r_3,G})$
      * $r_1,r_2,r_3\in\{1,2,...,n\}$ mutually different random idx + different from $i$
      * constant $F\in[0,2]$
    - crossover
      * \textbf{trial vector} $\boldsymbol{u}_{i,G+1}=(u_{i,G+1}^{(1)},u_{i,G+1}^{(2)},...,u_{i,G+1}^{(D)})$
        - chose random index $R\in\{1,2,...,D\}$
        - sample $p\sim U[0,1]$
        - define crossover constant $\text{CR}\in[0,1]$
        - $u_{i,G+1} = \begin{dcases}
          v_{i,G+1}^{(j)} & \text{if } p\le\text{CR}\text{ or } j=R \\
          x_{i,G}^{(j)} & \text{else}
         \end{dcases}$
    - selection: greedy, compare fitness of $\boldsymbol{u}_{i,G+1}$ with $\boldsymbol{x}_{i,G}$, pick better

@@

@@

@@

  \includegraphics[width=0.9\linewidth]{../code/export/plot_test_dhvs.png}  % scale figure, cf. https://tex.stackexchange.com/a/16584
  \caption{caption}

>>>>

(Taxonomy)
\citet[chap. 2]{survey_NN_interpretability} extend this distinction to a 3-dimensional taxonomy, allowing for better categorisation of neural networks (NN), a model class
that in its fully-connected feedforwad form is inherently non-interpretable.
% alternative formulation
%As fully-connected feedforward neural networks (NNs) are inherently non-interpretable models, the taxonomy introduced by \citet[chap. 2]{survey_NN_interpretability} proves
%particulary useful to categorise NN explanability methods along 3 dimensions:
- \textbf{Passive vs Active Approaches}
  * passive: post-hoc
  * active: changing architecture or training process to make model interpretable
- \textbf{Type of Explanations}
  * examples: providing examples of what leads to desired output
  * attribution: attributing effect on output for a specific feature
  * hidden semantics: examine what type of inputs specific neurons / layers pick up on
  * rules: logical rules, e.g. if-then, trees
- \textbf{Local vs Global Interpretability}
  * local: explanation based on individual samples
  * semi-local: explanation based on set of samples, e.g. grouped together by some criterion
  * global: explaining network as a whole

(Evaluation)
- objective evaluation difficult, as no clear definition of interpretability
- \citet[3]{DoshiVelez2017TowardsAR} propose a taxonomy to categorise possible evaluation methods
  * \textbf{application-grounded evaluation}
    - most rigorous method, also most expensive + time consuming
    - idea: evaluate ``interpretability model'' directly w.r.t. the task -> human experts evaluate the outcome
    - e.g.: model performing medical diagnosis on patients would be evaluated by based on doctors doing the same
  * \textbf{human-grounded metrics}
    - similar to application-grounded, but tries to simplify task (while preserving its essence) so that laypersons can do it
    - cheaper due to larger, less qualified subject pool
    - especially suitable if only general concepts of the tasks need to be validated
    - examples for evaluation set-up
      * binary forced choice: human evaluator chooses which of two explanations he finds better
      * forward simulation: evaluator must correctly simulate model output when presented model input + explanation % TODO: perhaps binary forced choice is already enough to include, this (and the third one is a bit awkward to explain why this is good for interpretability imo)
  * \textbf{functionally-grounded evaluation}
    - assess explanation quality via some formally defined proxy for interpretability
    - no human time nor cost required beyond initial formulation of proxy
    - e.g. if we already have an interpretable model class (e.g. identified via human-grounded evaluation), can then rank different models of that class based on proxy
    - main challenge: what proxies

@@

(evolutionary algorithms)
- another model-free class of optimisation algorithms, inspired by naturally occuring evolution
- based on a population, which iteratively (iteration = generation) generates $\lambda$ offspring via 3 operators (all sub-bullets from \citet[pp. 10-]{genetic_algos})
  * \textbf{reproduction}: pick an individual to reproduce with p proportional to its fitness
  * \textbf{crossover}
    - select 2 ``parents'' from pool of reproducing individuals
    - for a position $k$ in their HP configs, with certain p, swap their values after $k$, yielding 2 children
    % TODO: use example from book page 12
  * \textbf{mutation}: with certain p, change values of an individual, e.g. by adding Gaussian noise (Gaussian mutation) to real values, or flipping bit for binary values
- selection at end of each genertion: keep $\mu$ best individuals (based on some fitness function), either from only offspring (``$(\mu,\lambda)$-selection'')
  or (more commonly) from population + offspring (``$(\mu+\lambda)$-selection'', guaranteed to keep best individual)
- pro: conceptually simple + can handle even complex parameter spaces (continuous, discrete, hierarchical, etc.) given appropriate implementation of operators
all above starting at HPO from \citep[chap. 1.3]{feurer_hyperparameter_2019}
- prominent examples
  * CMA-ES
    - ``Covariance Matrix Adapation Evolution Strategy''
    - offspring generation exclusively via multivariate normal \citep[p. 8]{hansen2023cmaevolutionstrategytutorial}
      * mean = weighted average of previous generation
      * covariance = weighted covariance of previous generation, with weighting as for mean
    - weighing scheme done to sample in a way as to reproduce previously successful (i.e. selected) steps \citep[p. 11]{hansen2023cmaevolutionstrategytutorial}
  * differential evolution \citep[-]{differential_evolution}
    - init population randomly so that entire param space is covered
      * $D$-dim vectors $\boldsymbol{x}_{i,G}$ with $i=1,2,...,n$ for gen $G$
    - mutation
      * for each $\boldsymbol{x}_{i,G}$, generate \textbf{mutant vector} $\boldsymbol{v}_{i,G+1}=\boldsymbol{x}_{r_1,G}+F\cdot(\boldsymbol{x}_{r_2,G}-\boldsymbol{x}_{r_3,G})$
      * $r_1,r_2,r_3\in\{1,2,...,n\}$ mutually different random idx + different from $i$
      * constant $F\in[0,2]$
    - crossover
      * \textbf{trial vector} $\boldsymbol{u}_{i,G+1}=(u_{i,G+1}^{(1)},u_{i,G+1}^{(2)},...,u_{i,G+1}^{(D)})$
        - chose random index $R\in\{1,2,...,D\}$
        - sample $p\sim U[0,1]$
        - define crossover constant $\text{CR}\in[0,1]$
        - $u_{i,G+1} = \begin{dcases}
          v_{i,G+1}^{(j)} & \text{if } p\le\text{CR}\text{ or } j=R \\
          x_{i,G}^{(j)} & \text{else}
         \end{dcases}$
    - selection: greedy, compare fitness of $\boldsymbol{u}_{i,G+1}$ with $\boldsymbol{x}_{i,G}$, pick better

@@

  \includegraphics[width=\linewidth]{./architecture.png}  % scale figure, cf. https://tex.stackexchange.com/a/16584
  \caption{TODO: caption}
  \label{fig-nn-architecture}

@@

  \includegraphics[width=\linewidth]{./dataset_split.png}
  \caption{TODO: caption}
  \label{fig-dataset-split}

@@

  \includegraphics[width=0.9\linewidth]{../code/export/plot_test_dhvs.png}
  \caption{TODO: caption}
